---
title: "Individual assignment 5"
author: "Sami Gibson"
date: "2026-05-02"
format: pdf
execute:
  message: false
  warning: false
---
# 1. Set up

## Packages
```{r}
#| label: packages
# reads in tidyverse package from library 
library(tidyverse)
# reads in here package from library 
library(here)
# reads in janitor package from library 
library(janitor)
# reads in readxl package from library 
library(readxl)
# reads in snakecase package from library 
library(snakecase)
# reads in scales package from library 
library(scales)
# reads in patchwork package from library 
library(patchwork)
```

## Data
```{r}
#| label: data
#| 
# read in taxonomic information
taxon_list <- read_csv(here("data", "taxon_list.csv")) 

# read in aquatic invertebrates data
aq_ins <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                    sheet = "Aquatic Insects") 

# read in site information
sites <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                   sheet = "sites")

# read in water quality information
water_quality <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                           sheet = "Water Quality",
                           na = c("", "n/a", "N/A", "over", "173+")) 
```

# Cleaning
## `sites` object
```{r}
#| label: ncos-site-object

# creating new object from sites
ncos_sites <- sites |> 
  # filter to only include NCOS sites sampled four times a year
  filter(sector == "NCOS" & sampling_frequency == "Quarterly")
```
## Cleaning `taxon_list`
```{r}
#| label: cleaning-taxon-list

# creating new clean object from taxon_list
taxon_list_clean <- taxon_list |> 
  # converting taxon name to snake case (no spaces or capital letters,
  # only underscores)
  mutate(verbatim_name = to_any_case(verbatim_name, case = "snake"))
```
## Cleaning `water_quality`
```{r}
#| label: water-quality-cleaning

# creating new clean object from water quality
water_quality_clean <- water_quality |> 
  # clean column names
  clean_names() |> 
  # filtering join: only include sites in ncos_sites
  semi_join(ncos_sites, by = "site") |> 
  # create new column to join with later data frames
  unite("date_site", date, site, remove = TRUE) |> 
  # group by date_site column
  group_by(date_site) |> 
  # calculate median pH, salinity, DO
  summarize(med_ph = median(p_h, na.rm = TRUE),
            med_sal = median(salinity_ppt, na.rm = TRUE),
            med_do = median(dissolved_oxygen_mg_l, na.rm = TRUE)) |> 
  # ungroup data frame
  ungroup() |> 
  # filter out salinity outliers
  filter(date_site != "2022-08-12_NVBR") 
```
## Cleaning `aq_ins`
```{r}
#| label: cleaning-aq-ins

# creating new clean object from aquatic inverts
aq_ins_clean <- aq_ins |> 
  # clean column names
  clean_names() |> 
  # filtering join: only include sites occurring in ncos_sites
  semi_join(ncos_sites, by = c("site")) |> 
  # renaming columns
  rename(date = date_on_vial) |> 
  # select columns of interest
  select(date, sample_type, site,
         ostracod,
         copepod,
         hemiptera_corixidae_boatman,
         cladocera,
         nematode,
         diptera_ceratopogonidae, 
         annelida_oligochaete,
         diptera_chironomid,
         annelida_polychaete) |> 
  # filter to only include "filtered beaker" samples
  filter(sample_type %in% c("FB 250", "FB250")) |> 
  # convert the data frame to long format
  pivot_longer(cols = ostracod:annelida_polychaete,
               names_to = "taxon",
               values_to = "abundance") |> 
  # group by date, site, taxon
  group_by(date, site, taxon) |> 
  # calculate average abundance per liter (sum abundance divided by 7.5 liters)
  summarize(ave_lit = sum(abundance, na.rm = TRUE)/7.5) |> 
  # ungroup data frame
  ungroup() |> 
  # create a new column called `date_site` to join with water quality
  unite("date_site", date, site, remove = FALSE) |> 
  # join with water quality
  left_join(water_quality_clean, by = c("date_site")) |> 
  # join with taxon list
  left_join(taxon_list_clean, by = c("taxon" = "verbatim_name")) |> 
  # add full names of sites
  mutate(site_full = case_when(
    site == "NVBR" ~ "North of Venoco Bridge",
    site == "NEC" ~ "South of Venoco Bridge",
    site == "NMC" ~ "Main Channel",
    site == "NPB" ~ "South of Phelps Bridge",
    site == "NPB1" ~ "North of Phelps Bridge",
    site == "NPB2" ~ "Phelps Road",
    site == "NWP" ~  "West Pond",
    site == "NDC" ~ "Devereux Creek"
  )) |> 
  # setting factor levels for site
  mutate(site_full = fct_reorder(site_full, med_sal, .fun = "median", .na_rm = TRUE),
         site_full = fct_rev(site_full))
```

# Problems

## 1. Visualizing differences across sites in major taxon abundance
::: {.callout-note}
In class, we visualized the differences across sites in median (log transformed) copepod abundance.

In this section, you will create another figure to add to the multipanel plot of salinity and copepod abundance to show ostracod abundance.
:::
### a. Plan your figure
::: {.callout-note}
In part 1d, you should end up with a three panel figure. All three panels should have site on the x-axis, but each panel will have a different y-axis.

In part 1c, you will create one component of the three panel figure. Provide a list of the axes and geometries you will need to create a figure in 1c showing differences between sites in ostracod abundance.

x-axis:
y-axis:
geometry to represent average abundance/L:
geometry to represent medians:
:::

- x-axis: site (column: `site_full`)
- y-axis: log + 1 transformed average ostracod abundance/L
- geometry to represent average abundance/L: jittered circular open points (using: `geom_jitter()`)
- geometry to represent medians: larger diamond pink-filled points (using: `stat_summary()`)

### b. Wrangle your data
::: {.callout-note}
The aq_ins_clean data frame will need some further wrangling before you create the figure you described in part 1a.

In this section, create the data frame you will need to make your figure.

At the end of your code chunk, display 10 random rows from your data frame using slice_sample().
:::
```{r}
#| label: 1b wrangling data
#| 
# creating a new object ostracods from aq_ins_clean
ostracods <- aq_ins_clean |> 
  # filter to only include observations where taxon is ostracods
  filter(taxon == "ostracod") |> 
  # create new column that calculates log + 1 transform average abundance/liter
  mutate(log_ave_lit = log(ave_lit + 1))

# display 10 random rows from ostracod data frame
ostracods |> slice_sample(n = 10)
```

### c. Code your figure
::: {.callout-note}
Create the figure you outlined in part 1a in a way that mirrors the two-panel figure with which we ended class. For full marks:

make sure the x- and y-axes are labelled in a readable format
include a title that describes the taxonomic group being plotted
represent the median in a way that is visually distinct from each individual average abundance/L measurement with color, shape size, and shape type
use a theme that is different from the ggplot() default
:::
```{r}
#| label: 1c code your figure
#| fig-width: 10
#| fig-height: 6

# base layer: ggplot
ostracods_plot <- ggplot(data = ostracods, # assigns plot to object ostracods_plot 
                         # uses ostracods data frame
                         mapping = aes(x = site_full, # x-axis: site_full
                                       y = log_ave_lit)) + # y-axis: log_ave_lit
  # first layer: jittered points to show individual average abundance/L measurements
  geom_jitter(height = 0, # jitter height is 0
              width = 0.1, # jitter width is 0
              shape = 21, # jitter point shape is circle
              size = 2) + # jitter point size is 2
  # second layer: larger colored points to show median abundance at each site
  stat_summary(geom = "point", # uses point geometry
               fun = "median", # calculates the median
               shape = 18, # shape is diamond
               size = 4,# size is 4
               color = 'magenta3') + # color is pink
  # wraps long site names so x-axis labels are readable
  scale_x_discrete(labels = label_wrap(width = 15)) +
  labs(x = "Site", # relabel x-axis
       y = "log + 1 transformed average abundance/L", # relabel y-axis
       title = "Ostracods") + # title to describe what group is plotted
  # uses non-default theme
  theme_minimal()

# shows plot
ostracods_plot
```

### d. Create a multipanel figure
::: {.callout-note}
Combine the figure you made in 1c with the figures from class to create a multipanel figure using patchwork. Position the plots in the following arrangement:
:::
```{r}
#| label: 1d creating multipanel figure
#| fig-width: 14
#| fig-height: 12

# ADDING CLASS FIGURES

# creating new object from clean aquatic inverts 
copepods <- aq_ins_clean |> 
  # filter to only include copepods
  filter(taxon == "copepod") |> 
  # log + 1 transform mean abundance per liter
  mutate(log_ave_lit = log(ave_lit + 1)) 

# base layer
copepods <- ggplot(data = copepods,
                   # x-axis: site
                   # y-axis: mean abundance/L
                   mapping = aes(x = site_full,
                                 y = log_ave_lit)) +
  # first layer: points to represent observations
  geom_jitter(height = 0,
              shape = 21,
              width = 0.1) +
  # second layer: points to represent medians
  stat_summary(geom = "point",
               fun = "median",
               size = 4,
               color = "blue") +
  # wrapping x-axis text
  scale_x_discrete(labels = label_wrap(15)) +
  # relabelling x- and y-axis, adding title
  labs(x = "Site",
       y = "log + 1 transformed average abundance/L",
       title = "Copepods") +
  # different theme
  theme_bw()

# base layer
salinity <- ggplot(data = aq_ins_clean,
       # x-axis: site
       # y-axis: mean salinity
       mapping = aes(x = site_full,
                     y = med_sal)) +
  # first layer: salinity measurements at each survey
  geom_jitter(height = 0,
              width = 0.1,
              shape = 21) +
  # wrapping x-axis text
  scale_x_discrete(labels = label_wrap(width = 15)) +
  # second layer: points to represent means
  stat_summary(fun = median,
               color = "red",
               size = 1) +
  # relabelling x- and y-axes
  labs(x = "Site",
       y = "Median salinity (ppt)",
       title = "Site salinity") +
  # different theme
  theme_bw()


# combine salinity, copepod, and ostracod plots into one multipanel figure
salinity / (copepods | ostracods_plot)
```

### e. Interpret your figure
::: {.callout-note}
In 1-2 sentences each, describe:

differences between sites in copepod and ostracod abundance
how ostracod abundance may relate (or not) to site-level differences in salinity
Be specific using the visual components of the plot.
:::
**Differences between sites in copepod and ostracod abundance**: Copepod abundance varies by site, with the highest median value (indicated by large blue points) at South of Venoco Bridge and relatively low medians at the Main Channel and North of Phelps Bridge and the spread of individual observations (indicated by jittered black circles) shows substantial variability at some sites. In contrast, ostracod abundance is highest at South of Phelps Bridge (indicated by highest large pink diamond) with a wide spread of higher values, while most other sites (North of Venoco Bridge and South of Venoco Bridge) have medians near zero and tightly clustered low observations showing lower variability at the majority of the sites.

**How ostracod abundance may relate (or not) to site-level differences in salinity**: Ostracod abundance does not show a consistent relationship with salinity across sites as the highest ostracod median (shown by pink diamond) occurs at South of Phelps Bridge which has relatively low median salinity (shown by red circle), while higher salinity sites like North of Venoco Bridge have near-zero ostracod medians (shown by pink diamond). By looking at the lack of consistent pattern in the positions of the red salinity medians relative to the purple ostracod medians across sites I would infer that variables other than salinity may be influencing ostracod abundance.

## 2. Visualizing total abundance as a function of salinity
::: {.callout-note}
In this problem, you will create a figure visualizing log-transformed abundance as a function of salinity (measured in parts per thousand) to determine how salinity may influence invertebrate abundance (regardless of site).
:::
### a. Plan your figure
::: {.callout-note}
At the end of part 2c, you should have a figure:

depicting log-transformed average abundance/L on the y-axis and salinity on the x-axis
with different panels for each taxon in aq_ins_clean with free scales for both axes
with each taxon in its own color
Provide a list of values, geometries, and/or arguments that would be appropriate to make a figure matching this description.

x-axis:
y-axis:
color:
geometry:
facets:
:::

- x-axis: salinity (ppt)
- y-axis: log + 1 transformed average abundance/L (column: `log_ave_lit`)
- color: by taxon
- geometry: points (using `geom_point()`)
- facets: use function `facet_wrap()` by column `taxon` and arguement `scales = "free"`

### b. Wrangle your data
::: {.callout-note}
The aq_ins_clean data frame will need some further wrangling before you create the figure you described in part 1a.

In this section, manipulate the existing aq_ins_clean data frame to create new columns, modify existing columns, and/or select columns to create the figure you outlined in part 1a.

Additionally:

make sure each taxon is in title case
order taxon levels by maximum abundance
At the end of your code chunk, display 10 random rows from your data frame using slice_sample().
Revisit the directions from 2a and your outline to determine which columns you might keep from the data frame, which columns you might manipulate, and what new columns you might create.
:::
```{r}
#| label: 2b wrangle data

# create object sal_taxa from aq_ins_clean
sal_taxa <- aq_ins_clean |> 
  # remove rows that are missing median salinity values
  filter(!is.na(med_sal)) |> 
  # create column that calculates log-transformed abundance
  mutate(log_ave_lit = log(ave_lit + 1)) |>
  # mutates to title-case taxon names
  mutate(taxon = str_replace_all(taxon, "_", " "), taxon = str_to_title(taxon)) |>
  # order taxon levels by maximum average abundance per liter
  mutate(taxon = fct_reorder(taxon, ave_lit, .fun = max, .desc = TRUE)) |> 
  # keep only the columns needed for the figure
  select(date, site, site_full, taxon, ave_lit, log_ave_lit, med_sal)

# display 10 random rows from sal_taxa
sal_taxa |> slice_sample(n = 10)
```

### c. Code your figure
::: {.callout-note}
Create the figure you outlined in part 2a. For full marks:

make sure the x- and y-axes are labelled in a readable format
use a different theme from the default
use different colors from the default
take out the legend
:::
```{r}
#| label: 2c code your figure
#| fig-width: 10
#| fig-height: 8

# base layer: ggplot
sal_taxa_plot <- ggplot(data = sal_taxa, # assigns plot to object sal_taxa_plot 
                        # uses sal_taxa data frame
                        mapping = aes(x = med_sal, # x-axis: med_sal
                                      y = log_ave_lit, # y-axis: log_ave_lit
                                      color = taxon)) + # color by taxon
  # first layer: points show individual abundance measurement at each salinity value
  geom_point(shape = 16, # shape is circle
             size = 2, # size is 2
             alpha = 0.7) + # opacity is 0.7
  # separate panels for each taxon with free x- and y-axis scales
  facet_wrap(~ taxon, scales = "free") +
  # use non-default colors for points
  scale_color_viridis_d() +
  labs(x = "Median salinity (ppt)", # relabel x-axis
       y = "log + 1 transformed average abundance/L", # relabel y-axis
       # creates descriptive title
       title = "Invertebrate abundance as a function of salinity") +
  # uses non-default theme
  theme_minimal() +
  # remove redundant legend
  theme(legend.position = "none")

# shows plot
sal_taxa_plot
```

### d. Interpret your figure
::: {.callout-note}
In 3-5 sentences each, describe differences or similarities between taxa in relationships between abundance and salinity. Be specific using the visual components of the plot.
:::

**Differences between taxa in relationships between abundance and salinity**:
The relationship between abundance and salinity varies substantially across taxa which is seen by the different point distributions within each facet. Some taxa (including copepod, ostracod and cladocera) show higher abundances primarily at low salinity values (shown by clustered points with higher y-values near the left side of the x-axis), with abundances dropping off at higher salinities. Other taxa (such as Nematode and Annelida polychaete) are mostly concentrated near zero abundance across all salinities (indicated by points clustered around near-zero y-axis values), indicating little variation and no clear relationship.

**Similarities between taxa in relationships between abundance and salinity**:
Across some taxa (including: cladocera, ostracod, and copepod), there is a general pattern of higher abundances occurring at lower salinity values, shown by the clustering of higher y-values on the left side of many panels. Additionally, some taxa (including: Annelida oligochaete, Annelida polychaete, and Nematode) exhibit a large number of observations near zero abundance (points along the bottom of each panel), suggesting that low or absent abundance is most common regardless of salinity. Several taxa also show high variability at low salinity (wide vertical spread of points near low x-values), but much less variability at higher salinity, where points are more tightly clustered near zero. 

